Computing Entropies with Nested Sampling

### 2017-10-17 talk/ms.tex
Neil Lawrence, also references
Include figures from movie
Hard integrals
Finalising

@@

@@

@@

\begin{enumerate}
  \item<2-> Generate a `reference point' $\theta_{\rm ref}$ from $f$
  \item<3-> Do a Nested Sampling run with $f$ as ``prior'' and
            minus the distance to $\theta_{\rm ref}$ as
            ``likelihood''.
  \item<4-> Measure how many NS iterations were needed to make the
            distance to $\theta_{\rm ref}$ really small, and divide
            by $N$. That gives an
            unbiased estimate of the
            log-prob near $\theta_{\rm ref}$.
  \item<5-> Repeat steps 1--3 many times.
  \item<6-> Average the estimated log-probs, then apply corrections
            to convert to density.
\end{enumerate}

@@

% New slide
\begin{frame}
\frametitle{References I.}

On the connection between Shannon entropy and thermodynamic entropy,
see: \vspace{2em}

{\tiny Jaynes, Edwin T. ``Gibbs vs Boltzmann entropies.''
American Journal of Physics 33, no. 5 (1965): 391-398. \\

Brewer, Brendon J. ``Unscrambling the Second Law of Thermodynamics''
{\color{blue}
  \url{https://quillette.com/2016/01/28/unscrambling-the-second-law-of-thermodynamics}
}
} % tiny

@@

% New slide
\begin{frame}
\frametitle{Statements and questions}
Consider a problem with three mutually exclusive, exhaustive possibilities
\texttt{a}, \texttt{b}, and \texttt{c}.

\begin{center}
  \includegraphics[width=0.3\textwidth]{lattice1.pdf}
\end{center}

@@

>>>>

% New slide
\begin{frame}
\frametitle{Information Theory}

The fundamental theorem of information theory

% New slide
\begin{frame}
\frametitle{Information Theory}

The fundamental theorem of information theory
\begin{block}{Theorem}
If you take the log of a probability, it seems like you understand profound
truths.
\end{block}

@@

Heuristic: standard deviation quantifies uncertainty `horizontally',
entropy does it `vertically'.

@@

% New slide
\begin{frame}
\frametitle{Hard integrals}
Hard integrals with nuisance parameters
$\boldsymbol{\eta}$, interesting parameter(s)
$\boldsymbol{\phi}$
\vspace{0.5em}

\begin{center}
\includegraphics[width=0.9\textwidth]{hard-integrals.png}
\end{center}

@@

\begin{enumerate}
  \item<2-> Generate a `reference point' $\theta_{\rm ref}$ from $f$
  \item<3-> Do a Nested Sampling run with $f$ as ``prior'' and
            minus the distance to $\theta_{\rm ref}$ as
            ``likelihood''.
  \item<4-> Measure how many NS iterations were needed to make the
            distance to $\theta_{\rm ref}$ really small, and divide
            by $N$. That gives an
            unbiased estimate of the
            log-prob near $\theta_{\rm ref}$.
  \item<5-> Repeat steps 1--3 many times.
  \item<6-> Average the estimated log-probs, then apply corrections
            to convert to density.
\end{enumerate}
\end{frame}

% New slide
\begin{frame}
\frametitle{My algorithm}

\begin{center}
\includegraphics[width=0.7\textwidth]{movie/image000001.png}
\end{center}

@@

% New slide
\begin{frame}
\frametitle{My algorithm}

\begin{center}
\includegraphics[width=0.7\textwidth]{movie/image000021.png}
\end{center}

@@

% New slide
\begin{frame}
\frametitle{My algorithm}

\begin{center}
\includegraphics[width=0.7\textwidth]{movie/image000041.png}
\end{center}

@@

% New slide
\begin{frame}
\frametitle{Movie}

Play movie.mkv

(It's also on YouTube)

% New slide
\begin{frame}
\frametitle{Toy experimental design example}
Two observing strategies --- even and uneven.
Which is better for measuring a period?

\begin{center}
\includegraphics[width=0.8\textwidth]{../paper/figures/sinewave.pdf}
\end{center}

% New slide
\begin{frame}
\frametitle{Specifics}
Let $\tau = \log_{10}(\textnormal{period})$.\vspace{0.5em}

I knew $H(\tau)$ because I chose
the prior. I used the algorithm to estimate $H(\tau | d)$,
so marginal posteriors
were the distributions whose entropies I estimated\footnote{If you only
care about one parameter, define your distance function in terms of that
parameter only!}.\vspace{0.5em}

I then computed the mutual information
\vspace{0.5em}

\begin{align}
I(\tau; d) &= H(\tau) - H(\tau | d)
\end{align}

% New slide
\begin{frame}
\frametitle{Result}

\begin{align}
I_{\rm even}   &= 5.441 \pm 0.038 \textnormal{ nats}\\
I_{\rm uneven} &= 5.398 \pm 0.038 \textnormal{ nats}
\end{align}

i.e., any difference is trivial.

% New slide
\begin{frame}
\frametitle{In practice...}

When a period is short relative to observations, you can get a multimodal
posterior pdf\footnote{e.g., see Larry Bretthorst's work connecting
the posterior pdf to the periodogram.},
and `learn a lot' by ruling out most of the space, but still
having many peaks.\vspace{0.5em}

I did not investigate this aspect of the problem --- I assumed long periods.

% New slide
\begin{frame}
\frametitle{Paper/software/endorsement}

{\color{blue}\url{http://www.mdpi.com/1099-4300/19/8/422}
\url{https://github.com/eggplantbren/InfoNest}}
\vspace{0.5cm}

\begin{center}
\includegraphics[width=0.7\textwidth]{lawrence.png}
\end{center}

%% New slide
%\begin{frame}
%\frametitle{References I.}

%On the connection between Shannon entropy and thermodynamic entropy,
%see: \vspace{2em}

%{\tiny Jaynes, Edwin T. ``Gibbs vs Boltzmann entropies.''
%American Journal of Physics 33, no. 5 (1965): 391-398. \\

%Brewer, Brendon J. ``Unscrambling the Second Law of Thermodynamics''
%{\color{blue}
%  \url{https://quillette.com/2016/01/28/unscrambling-the-second-law-of-thermodynamics}
%}
%} % tiny

%\end{frame}

%% New slide
%\begin{frame}
%\frametitle{Statements and questions}
%Consider a problem with three mutually exclusive, exhaustive possibilities
%\texttt{a}, \texttt{b}, and \texttt{c}.

%\begin{center}
%  \includegraphics[width=0.3\textwidth]{lattice1.pdf}
%\end{center}

%\end{frame}
